Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-13 manuscript.tex
Update on Overleaf.

% a simpler annotation system:
\usepackage{soulutf8}
\usepackage{color}
\newcommand{\note}[1]{{\sethlcolor{yellow}\hl{#1}}}
\usepackage[margins]{trackchanges}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONTEXT
Neural activity is directly influenced by our immediate environment and by internal states and is structured in order to generate motor actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as exemplified by the recordings observed in neurobiological experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality to denote the existence of a structure in neural activity. In this respect, it is known that neural activity may be related to specific measurable variables, and since Galvani's experiments, we know that an electrical activity applied on muscular nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A central and well-studied way of communication between neurons is specific electrochemical events called action potentials, or spikes, which were first discovered by \href{https://en.wikipedia.org/wiki/Edgar_Adrian}{Lord Adrian}~\citep{adrian_impulses_1926}. In the scope of this article, we focus mainly on these spikes, which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude, i.e. they are prototypical, 
``\href{https://en.wikipedia.org/wiki/All-or-none_law}{all-or-none}''. \change[Antoine]{Typically, these are received by a neuron from other afferent neurons on the arborized ``input'' dendrite, integrated in its soma, and possibly further propagated along its ``output'' axon to reach efferent neurons.}{Typically, a postsynaptic neuron receives incoming spikes from other afferent neurons on the arborized ``input'' dendrite. Once integrated by the dendritic tree and the soma of the neuron, these spikes result in a modification of its membrane potential that possibly leads to the emission of an action potential along its ``output'' axon to reach efferent neurons.}
The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code. Indeed, most neurophysiologists use the temporal evolution of the firing rate (for instance, as computed as the average occurrence of spikes in small temporal windows of about $100~\ms$) in order to characterize the dynamical activity of neurons. This may be extended by computing different statistics on each neuron's sequence of spikes~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role and that neurons may be synchrony detectors as well as integrators~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. %Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. 
In comparison to a classical analog vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of representations, that rely on the firing rate, by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a focus on the state-of-the-art of the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us to make sense of the world through luminous signals, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed by Lord Adrian when recording the activity the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is remarkably a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.75\textwidth]{figures/visual-latency.pdf}
% thorpe
% \includegraphics[width=.7\textwidth]{figures/visual-latency-estimate.jpg}%-estimate.jpg}
\caption{{\bf Latency of the different processing steps along the visual pathway.} Though this system is highly inter-connected, one can follow the sequence of activations whenever an image (here a yellow star) is flashed in front of the eyes. Different areas are schematically represented by ellipses and arrows denote the fastest feed-forward activation, ordered with respect to their activation latency. In order, the retina is first activated ($20-40~\ms$) then the thalamus and the primary visual cortex (V1, $60-80~\ms$). Visual information used for object recognition follows the temporal lobe to reach the infero-temporal area (IT, $120-140~\ms$), and then the prefrontal cortex (PFC) which modulates decision making and the Motor cortex (MC, $140-190~\ms$) which mediates a motor action which is then relayed through the spinal cord to trigger finger muscles, with latencies about $200-350~\ms$.}\label{fig:thorpe}
\end{figure}
%----------------------------%
In particular, there have been some remarkable findings when studying the dynamics of vision. For instance, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. His experiments consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges with a very short latency in neural activity. These results have been extended to several species, including primates. Different experimental protocols have shown for example that the motor response could be extremely fast (of the order of $120~\ms$) when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}. This fast processing correlates with the surprising experiments of fast serial detection, which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal~\citep{keysers_speed_2001}. As expected, the performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only $14~\ms$ per image.

Although surprising, this speed of the visual cortex in primates is compatible with the latencies that are recorded at the neuro-physiological level. The rapid propagation of the visual information in the thalamus, then in the primary visual cortex takes about $45~\ms$ in the macaque~\citep{schmolesky_signal_1998} and about $60~\ms$ in humans~\citep{vanni_coinciding_2001}. This functioning of visual processing as a forward pass is most prominent in fast processing (see Figure~\ref{fig:thorpe}), and can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}. An important consequence of this speed of processing of vision is that it implies that processing is carried out using only very few spikes per layer. As a comparison, the latencies in macaque monkey are about: Retina, $20-40~\ms$; V1, $40-60~\ms$; IT, $80-100~\ms$; MC, $140-190~\ms$; to finger muscles, $180-260~\ms$. Indeed, we may consider that a behavioral response occurring in only $200~\ms$ would involve about ten processing stages along the ``forward'' pathways of the visual system. Such models were indeed efficiently modeled in feed-forward models trained with back-propagation~\citep{serre_feedforward_2007,jeremie_ultrafast_2022}. At the same time, it has been demonstrated that one spike requires a significant amount of time (about $10~\ms$) to be conducted from one layer to the next~\citep{nowak_timing_1997,thorpe_seeking_2001}.  This figure is inspired by a similar schematics done for monkeys in~\citep{thorpe_seeking_2001}. As a consequence, these results suggest that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How precise spike timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% reliability of spike timings
Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons which hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: while the first spike is aligned to stimulus' time, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

% retina
How may information about the sensory world be encoded in the spiking pattern?  At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{\textbf{Reproducibility of the spiking response of a neuron. }The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, it highlights the consequence of using the same \emph{frozen} noise at each repetition and highlight the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of the results from~\citep{mainen_reliability_1995}  using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%

>>>>

% a simpler annotation system:
\usepackage{soulutf8}
\usepackage{color}
\newcommand{\note}[1]{{\sethlcolor{yellow}\hl{#1}}}
\usepackage{trackchanges}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction: Importance of precise spike timings in the brain}\label{sec:time}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Is there a neural code?}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% CONTEXT
Neural activity is directly influenced by our immediate environment and by internal states and is structured in order to generate motor actions. The efficiency of these actions is key for survival, which is the sole determinant in the light of natural selection. A central question of modern neuroscience is to better understand the essence of neural activity, as exemplified by the recordings observed in neurobiological experiments. One sometimes uses the expression ``decoding the neural code'', although this implies the existence of a code, i.e. an explicit representation within the neural activity. Nevertheless, we will use this terminology in all generality to denote the existence of a structure in neural activity. In this respect, it is known that neural activity may be related to specific measurable variables, and since Galvani's experiments, we know that an electrical activity applied on muscular nerves can cause the stretching of a frog's limb (for a review, see~\citep{piccolino_luigi_1997}). A central and well-studied way of communication between neurons is specific electrochemical events called action potentials, or spikes, which were first discovered at the beginning of the $XX^{th}$ century~\citep{adrian_impulses_1926}.
In the scope of this article, we focus mainly on these spikes, which are brief (about one millisecond) impulses that propagate along the axons of neurons. First, these have the particular property of being essentially binary in their amplitude, i.e. they are prototypical, 
``\href{https://en.wikipedia.org/wiki/All-or-none_law}{all-or-none}''. \change[Antoine]{Typically, these are received by a neuron from other afferent neurons on the arborized ``input'' dendrite, integrated in its soma, and possibly further propagated along its ``output'' axon to reach efferent neurons.}{Typically, a postsynaptic neuron receives incoming spikes from other afferent neurons on the arborized ``input'' dendrite. The integration of these spikes by the dendritic tree and the soma of the postsynaptic neuron results in the modification of its membrane potential that possibly leads to the emission of an action potential along its ``output'' axon to reach efferent neurons.}
The frequency at which these spikes are emitted has been shown to be roughly commensurate with the stretch of the frog's limb~\citep{adrian_impulses_1926}, and this variable (measured in spikes per second, or Hertz) is as of today considered to be the main constituent of the neural code \note{change that and add a more wide intro about other possible constituent of the neural code}. \change[Antoine]{Indded, neurophysiologists typically}{Most neurophysiologists} use the temporal evolution of the firing rate (for instance, as computed as the average occurrence of spikes in small temporal windows of about $100~\ms$) in order to characterize the dynamical activity of neurons. This may be extended by computing different statistics on each neuron's sequence of spikes~\citep{perkel_neuronal_1967} but also the dependence across neurons~\citep{perkel_neuronal_1967-1}.

% GAP
However, computational neuroscience models have suggested that the precise timing within a sequence of spikes may play a crucial role and that neurons \change[Antoine]{may be as well integrators than synchrony detectors}{may be synchrony detectors as well as integrators}~\citep{abeles_role_1982}. We will investigate this very hypothesis in this review. %Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma~\citep{paugam-moisy_computing_2012}. 
In comparison to a classical \change[Antoine]{analogical}{analog} vector of inputs, this \emph{event-based representation} observed in the neural code is essential in understanding information processing~\citep{carr_processing_1993}. For instance, it expands the capabilities of representations \change[Antoine]{(which rely on the firing rate)}{that rely on the firing rate} by considering a representation based on repetitions of spiking motifs at precise times of occurrence. This hypothesis is directly inspired by neurobiological observations and expands the capabilities of analog representations (which are based on spiking firing rates) by considering a representation based on instances of spiking motifs at precise times of occurrence. Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004,dilorenzo_spike_2013}, highlight the importance of the temporal aspect of the neural code and further suggest the existence of precise spatio-temporal spiking motifs in the input which excites neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing~\citep{roy_towards_2019}, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Crucially, validating this hypothesis would also be crucial in our understanding of neural processes. %
\subsection{Dynamics of vision and consequences on the neural code}%Precise timing in the neural code enables efficient vision}
Let us start with a focus on the state-of-the-art of the role of dynamics in vision. Broadly speaking, vision is the set of processes that allow us \change[Antoine]{to make sense of the luminous world}{to make sense of the world through luminous signals}, and is an intensively studied field in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between the topology of external visual space and its representation on the retina, called retinotopy. The origins of this neuroscientific question can be found in the first experiments from Pierre Flourens which, using lesions in animals, demonstrated the relationship between visual sensations and activity in the cerebral cortex~\citep{flourens_recherches_1842,pearce_marie-jean-pierre_2009}. This was also observed \remove[Antoine]{by Lord Adrian} when recording the activity of the frog's visual system~\citep{adrian_impulses_1926}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as the visual orientation or motion of elementary contours. For a large proportion of neurons, there is \change[Antoine]{remarkably a monotonic}{a remarkably monotonic} relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

%----------------------------%
\begin{figure}
\centering
\includegraphics[width=.75\textwidth]{figures/visual-latency.pdf}
% thorpe
% \includegraphics[width=.7\textwidth]{figures/visual-latency-estimate.jpg}%-estimate.jpg}
\caption{{\bf Latency of the different processing steps along the visual pathway.} Though this system is highly inter-connected, one can follow the sequence of activations whenever an image (here a yellow star) is flashed in front of the eyes. Different areas are schematically represented by ellipses and arrows denote the fastest feed-forward activation, ordered with respect to their activation latency. In order, the retina is first activated ($20-40~\ms$) then the thalamus and the primary visual cortex (V1, $60-80~\ms$). Visual information used for object recognition follows the temporal lobe to reach the infero-temporal area (IT, $120-140~\ms$), and then the prefrontal cortex (PFC) which modulates decision making and the Motor cortex (MC, $140-190~\ms$) which mediates a motor action which is then relayed through the spinal cord to trigger finger muscles, with latencies about $200-350~\ms$.}\label{fig:thorpe}
\end{figure}
\note{Je changerai cette figure car nous parlons du système visuel des primates non humains}
%----------------------------%
In particular, there have been some remarkable findings when studying the dynamics of vision. For instance, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. \change[Antoine]{His}{Their} experiments consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges with a very short latency in neural activity. These results have been extended to several species, including primates. Different experimental protocols have shown for example that the motor response could be extremely fast (of the order of $120~\ms$) when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}. This fast processing correlates with the surprising experiments of fast serial detection, which consists in presenting a fast succession of different images and decoding via the EEG if the observer can detect for example the presence of an animal~\citep{keysers_speed_2001}. As expected, the performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only $14~\ms$ per image.

Although surprising, this speed of the visual cortex in primates is compatible with the latencies that are recorded at the neuro-physiological level. The rapid propagation of the visual information in the thalamus, then in the primary visual cortex takes about $45~\ms$ in the macaque~\citep{schmolesky_signal_1998} and about $60~\ms$ in humans~\citep{vanni_coinciding_2001}. This functioning of visual processing as a forward pass is most prominent in fast processing (see Figure~\ref{fig:thorpe}), and can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}. An important consequence of this speed of processing of vision is that it implies that processing is carried out using only very few spikes per layer. As a comparison, the latencies in macaque monkeys are about: Retina, $20-40~\ms$; V1, $40-60~\ms$; IT, $80-100~\ms$; MC, $140-190~\ms$; to finger muscles, $180-260~\ms$. Indeed, we may consider that a behavioral response occurring in only $200~\ms$ would involve about ten processing stages along the ``forward'' pathways of the visual system. Such models were indeed efficiently modeled in feed-forward models trained with back-propagation~\citep{serre_feedforward_2007,jeremie_ultrafast_2022}. At the same time, it has been demonstrated that one spike requires a significant amount of time (about $10~\ms$) to be conducted from one layer to the next~\citep{nowak_timing_1997,thorpe_seeking_2001}.  This figure is inspired by similar schematics done for monkeys in~\citep{thorpe_seeking_2001}. As a consequence, these results suggest that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{How precise spike timing may encode vectors of real values}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% reliability of spike timings
Sensory data is most often represented by continuous values, such as the energy produced by a flow of photons that hits the different photoreceptors of the retina. How may such information be encoded in neuronal activity? The analysis of generic raster plots reveals particular traits that hint at the role of a precise timing. For instance, the firing rate of cortical cells in awake monkeys is highly irregular~\citep{softky_highly_1993}, which makes it at first sight inconsistent with the temporal integration of firing rate. Remarkably, it has been observed that the response of a neuron in a cortical slice to a current step could be highly non-reproducible: while the first spike is aligned to stimulus' time, the subsequent spike times tend to diffuse for independent repetitions of the stimulation~\citep{bryant_spike_1976,mainen_reliability_1995} (see Figure~\ref{fig:mainen}). However, if that same neuron is now driven by a \emph{frozen} noise, that is, a highly dynamic signal which is repeatedly replayed from trial to trial, then the output spikes are highly reproducible (for a review, see~\citep{ermentrout_reliability_2008}). This is consistent with the differential role of different stimulus frequencies (for instance the gamma range around $80~\si{\Hz}$) on the reliability of spike timing reported in~\citep{nowak_influence_1997} : ``we found that, as expected given the resistive and capacitive properties of cortical neurons, low frequencies have a larger effect on the membrane potential of cortical neurons than do higher frequencies. However, increasing the amount of gamma range fluctuations in a stimulus leads to more precise timing of action potentials.''

% retina
How may information about the sensory world be encoded in the spiking pattern? At the level of the retina, it has been shown that a coding of luminance values in the image using the timing of the spikes may be at work~\citep{gollisch_rapid_2008}. \note{here we have to make the choice of studying only vertebrates, otherwise there is a huge review to add on insects. Gollisch and Meister is the fly retina? Maybe find another ref} In particular, these results show that the response of ganglion cells to the visual gratings projected on the retina could be encoded in the latency of the response and not only in the frequency of the discharge, as it is often assumed. These results have been extended to natural images and show a qualitatively similar behavior. The authors' conclusion is that the precise spiking latency of the neurons encodes the spatial features of the image. Interestingly, such a precise latency mechanism may underlie some visual illusions, e.g., the false color illusion in the \href{https://michaelbach.de/ot/col-Benham/index.html}{Benham Top} based on center-surround interactions in the parvocellular pathway~\citep{kenyon_theory_2004}. This evidence found in the retina can be extended to other areas such as the visual cortex.
%----------------------------%
% C_MainenSejnowski1995_Perrinet.ipynb
\begin{figure}
\centering
\includegraphics[width=\textwidth]{figures/replicating_MainenSejnowski1995.pdf} 
\caption{\textbf{Reproducibility of the spiking response of a neuron. }The timing of the spikes produced following the repetition of a step stimulus is less reproducible than that to a noisy stimulus. The stimulus current value over time for a step stimulus (top left) and for a noisy one (top right). Trial repetitions of a Leaky Integrate-and-Fire neuron stimulated by the stimulus on the upper row (middle row). Membrane potential is represented by dark blue color when light yellow colors when depolarized) and quantified by . The average firing rate across trials (lower row). While this seems paradoxical at first sight, it highlights the consequence of using the same \emph{frozen} noise at each repetition and highlights the highly reproducible pattern of spikes when it is driven by a highly dynamic input. See this~\href{https://github.com/laurentperrinet/2022_UE-neurosciences-computationnelles/blob/master/C_MainenSejnowski1995_Perrinet.ipynb}{notebook} for a replication of the results from~\citep{mainen_reliability_1995}  using a simple LIF model.}\label{fig:mainen}
\end{figure}
%----------------------------%
